---
title: "Week 9 - Homework"
author: "STAT 420, Summer 2021, D. Unger"
date: ''
output:
  html_document: 
    theme: readable
    toc: yes
  pdf_document: default
urlcolor: cyan
---


# Directions

Students are encouraged to work together on homework. However, sharing, copying or providing any part of a homework solution or code is an infraction of the University's rules on Academic Integrity. Any violation will be punished as severely as possible.

- Be sure to remove this section if you use this `.Rmd` file as a template.
- You may leave the questions in your final document.

***

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```

## Exercise 1 (`longley` Macroeconomic Data)

The built-in dataset `longley` contains macroeconomic data for predicting employment. We will attempt to model the `Employed` variable.

```{r, eval = FALSE}
View(longley)
?longley
```

**(a)** What is the largest correlation between any pair of predictors in the dataset?

```{r}
cor(longley)
```

The largest correlation between any pair of predictors in the dataset is 0.9952735 between Year and GNP.

**(b)** Fit a model with `Employed` as the response and the remaining variables as predictors. Calculate and report the variance inflation factor (VIF) for each of the predictors. Which variable has the largest VIF? Do any of the VIFs suggest multicollinearity?

```{r}
library(faraway)
model = lm(Employed ~ ., longley)
vif(model)
```

GNP has the largest VIF of 1788.51348. A VIF greater than 5 suggest collinearity. This means that GNP.deflator, GNP, Unemployed, Population, and Year all suggest collinearity of predictors in this model.

**(c)** What proportion of the observed variation in `Population` is explained by a linear relationship with the other predictors?

```{r}
model_pop = lm(Population ~ . - Employed, longley)
summary(model_pop)$r.squared
```

99.74947% of the variation in Population can be explained by a linear relationship with other predictors.

**(d)** Calculate the partial correlation coefficient for `Population` and `Employed` **with the effects of the other predictors removed**.

```{r}
model = lm(Employed ~ . - Population, longley)
cor(resid(model_pop), resid(model))
```

**(e)** Fit a new model with `Employed` as the response and the predictors from the model in **(b)** that were significant. (Use $\alpha = 0.05$.) Calculate and report the variance inflation factor for each of the predictors. Which variable has the largest VIF? Do any of the VIFs suggest multicollinearity?

```{r}
model = lm(Employed ~ ., longley)
summary(model)

model_small = lm(Employed ~ Unemployed + Armed.Forces + Year, longley)
vif(model_small)
```

Year has the largest VIF of 3.890861. Because a VIF greater than 5 suggest collinearity, none of the variables in this smaller model suggest collinearity.

**(f)** Use an $F$-test to compare the models in parts **(b)** and **(e)**. Report the following:

- The null hypothesis

$\beta_{GNP} = \beta_{GNP.deflator} = \beta_{Population} = 0$

- The test statistic

```{r}
model = lm(Employed ~ ., longley)
model_small = lm(Employed ~ Unemployed + Armed.Forces + Year, longley)

anova(model_small, model)$F[2]
```

- The distribution of the test statistic under the null hypothesis

```{r}
anova(model_small, model)
```

9 and 3 degrees of freedom.

- The p-value

```{r}
anova(model_small, model)["Pr(>F)"]
```

- A decision

Because the P-Value is greater than $\alpha = 0.05$, we fail to reject the null hypothesis.

- Which model you prefer, **(b)** or **(e)**

Because we fail to reject the null hypothesis, we prefer the smaller model.

**(g)** Check the assumptions of the model chosen in part **(f)**. Do any assumptions appear to be violated?

```{r, echo = FALSE}
plot_fitted_resid = function(model, pointcol = "dodgerblue", linecol = "darkorange") {
  plot(fitted(model), resid(model), 
       col = pointcol, pch = 20, cex = 1.5,
       xlab = "Fitted", ylab = "Residuals")
  abline(h = 0, col = linecol, lwd = 2)
}

plot_qq = function(model, pointcol = "dodgerblue", linecol = "darkorange") {
  qqnorm(resid(model), col = pointcol, pch = 20, cex = 1.5)
  qqline(resid(model), col = linecol, lwd = 2)
}

par(mfrow = c(1, 2))
plot_fitted_resid(model_small)
plot_qq(model_small)
```

With the fitted versus residuals plot, it appears that the points are centered at 0. There are a few points that have higher variance, but in general, at each fitted value, variance is equal. In the Q-Q plot, we notice that the tails are slightly fat. This means that normality may be in question.

***

## Exercise 2 (`Credit` Data)

For this exercise, use the `Credit` data from the `ISLR` package. Use the following code to remove the `ID` variable which is not useful for modeling.

```{r}
library(ISLR)
data(Credit)
Credit = subset(Credit, select = -c(ID))
```

Use `?Credit` to learn about this dataset.

**(a)** Find a "good" model for `balance` using the available predictors. Use any methods seen in class except transformations of the response. The model should:

- Reach a LOOCV-RMSE below `140`
- Obtain an adjusted $R^2$ above `0.90`
- Fail to reject the Breusch-Pagan test with an $\alpha$ of $0.01$
- Use fewer than 10 $\beta$ parameters

Store your model in a variable called `mod_a`. Run the two given chunks to verify your model meets the requested criteria. If you cannot find a model that meets all criteria, partial credit will be given for meeting at least some of the criteria.

```{r}
model = lm(Balance ~ ., Credit)

n = length(resid(model))

step(model, direction = "backward", k = log(n))

mod_a = lm(Balance ~ log(Income) + Married + I(Limit^3) + Rating + log(Cards) + I(Age^3) + Student + Gender, Credit)
```

After performing a backwards BIC test, I added a few predictors and transformed some of the predictors to achieve the above tests.

```{r, message = FALSE, warning = FALSE}
library(lmtest)

get_bp_decision = function(model, alpha) {
  decide = unname(bptest(model)$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_sw_decision = function(model, alpha) {
  decide = unname(shapiro.test(resid(model))$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_num_params = function(model) {
  length(coef(model))
}

get_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

get_adj_r2 = function(model) {
  summary(model)$adj.r.squared
}
```

```{r, eval = TRUE}
get_loocv_rmse(mod_a)
get_adj_r2(mod_a)
get_bp_decision(mod_a, alpha = 0.01)
get_num_params(mod_a)
```

**(b)** Find another "good" model for `balance` using the available predictors. Use any methods seen in class except transformations of the response. The model should:

- Reach a LOOCV-RMSE below `130`
- Obtain an adjusted $R^2$ above `0.85`
- Fail to reject the Shapiro-Wilk test with an $\alpha$ of $0.01$
- Use fewer than 25 $\beta$ parameters

Store your model in a variable called `mod_b`. Run the two given chunks to verify your model meets the requested criteria. If you cannot find a model that meets all criteria, partial credit will be given for meeting at least some of the criteria.

```{r}
model = lm(Balance ~ ., Credit)

n = length(resid(model))

step(model, direction = "backward", k = log(n))

mod_b = lm(Balance ~ (log(Income) + Limit + Cards + Student) ^ 4, Credit)
```

Again, I took the "best" model from the backwards BIC test. Then, I attempted to modify the predictors to meet all of the criteria.

```{r, message = FALSE, warning = FALSE}
library(lmtest)

get_bp_decision = function(model, alpha) {
  decide = unname(bptest(model)$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_sw_decision = function(model, alpha) {
  decide = unname(shapiro.test(resid(model))$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_num_params = function(model) {
  length(coef(model))
}

get_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

get_adj_r2 = function(model) {
  summary(model)$adj.r.squared
}
```

```{r, eval = TRUE}
get_loocv_rmse(mod_b)
get_adj_r2(mod_b)
get_sw_decision(mod_b, alpha = 0.01)
get_num_params(mod_b)
```

***

## Exercise 3 (`Sacramento` Housing Data)

For this exercise, use the `Sacramento` data from the `caret` package. Use the following code to perform some preprocessing of the data.

```{r}
library(caret)
library(ggplot2)
data(Sacramento)
sac_data = Sacramento
sac_data$limits = factor(ifelse(sac_data$city == "SACRAMENTO", "in", "out"))
sac_data = subset(sac_data, select = -c(city, zip))
```

Instead of using the `city` or `zip` variables that exist in the dataset, we will simply create a variable (`limits`) indicating whether or not a house is technically within the city limits of Sacramento. (We do this because they would both be factor variables with a **large** number of levels. This is a choice that is made due to laziness, not necessarily because it is justified. Think about what issues these variables might cause.)

Use `?Sacramento` to learn more about this dataset.

A plot of longitude versus latitude gives us a sense of where the city limits are.

```{r}
qplot(y = longitude, x = latitude, data = sac_data,
      col = limits, main = "Sacramento City Limits ")
```

After these modifications, we test-train split the data.

```{r}
set.seed(420)
sac_trn_idx  = sample(nrow(sac_data), size = trunc(0.80 * nrow(sac_data)))
sac_trn_data = sac_data[sac_trn_idx, ]
sac_tst_data = sac_data[-sac_trn_idx, ]
```

The training data should be used for all model fitting. Our goal is to find a model that is useful for predicting home prices.

**(a)** Find a "good" model for `price`. Use any methods seen in class. The model should reach a LOOCV-RMSE below 77,500 in the training data. Do not use any transformations of the response variable.

```{r}
model = lm(price ~ ., sac_trn_data)

n = length(resid(model))

step(model, direction = "backward", k = log(n))

model = lm(formula = price ~ beds + sqft + longitude + latitude, data = sac_trn_data)

sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
```

Here, I used the backward BIC to find the "best" model. From there, I simply added the latitude predictor to achieve a LOOCV-RMSE below 77,500.

**(b)** Is a model that achieves a LOOCV-RMSE below 77,500 useful in this case? That is, is an average error of 77,500 low enough when predicting home prices? To further investigate, use the held-out test data and your model from part **(a)** to do two things:

- Calculate the average percent error:
\[
\frac{1}{n}\sum_i\frac{|\text{predicted}_i - \text{actual}_i|}{\text{predicted}_i} \times 100
\]

```{r}
mean(abs((predict(model, sac_tst_data)) - sac_tst_data$price) / (predict(model, sac_tst_data))) * 100
```

- Plot the predicted versus the actual values and add the line $y = x$.

```{r}
plot(sac_tst_data$price, predict(model, sac_tst_data), xlab = "Actual Home Price", ylab = "Predicted Home Price", main = "Predicted versus Actual Home Price", col = "darkorange")
abline(0, 1, lwd = 2, col = "purple")
```


Based on all of this information, argue whether or not this model is useful.

For lower home prices, the model is a better predictor than for higher home prices. However, with an average error of roughly 25%, it is easy to argue that the model is not useful. Perhaps, it would be more useful to use 2 models, one for lower home prices and one for higher home prices.

***

## Exercise 4 (Does It Work?)

In this exercise, we will investigate how well backwards AIC and BIC actually perform. For either to be "working" correctly, they should result in a low number of both **false positives** and **false negatives**. In model selection,

- **False Positive**, FP: Incorrectly including a variable in the model. Including a *non-significant* variable
- **False Negative**, FN: Incorrectly excluding a variable in the model. Excluding a *significant* variable

Consider the **true** model

\[
Y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_4 + \beta_5 x_5 + \beta_6 x_6 + \beta_7 x_7 + \beta_8 x_8 + \beta_9 x_9 + \beta_{10} x_{10} + \epsilon
\]

where $\epsilon \sim N(0, \sigma^2 = 4)$. The true values of the $\beta$ parameters are given in the `R` code below.

```{r}
beta_0  = 1
beta_1  = -1
beta_2  = 2
beta_3  = -2
beta_4  = 1
beta_5  = 1
beta_6  = 0
beta_7  = 0
beta_8  = 0
beta_9  = 0
beta_10 = 0
sigma = 2
```

Then, as we have specified them, some variables are significant, and some are not. We store their names in `R` variables for use later.

```{r}
not_sig  = c("x_6", "x_7", "x_8", "x_9", "x_10")
signif = c("x_1", "x_2", "x_3", "x_4", "x_5")
```

We now simulate values for these `x` variables, which we will use throughout part **(a)**.

```{r}
set.seed(420)
n = 100
x_1  = runif(n, 0, 10)
x_2  = runif(n, 0, 10)
x_3  = runif(n, 0, 10)
x_4  = runif(n, 0, 10)
x_5  = runif(n, 0, 10)
x_6  = runif(n, 0, 10)
x_7  = runif(n, 0, 10)
x_8  = runif(n, 0, 10)
x_9  = runif(n, 0, 10)
x_10 = runif(n, 0, 10)
```

We then combine these into a data frame and simulate `y` according to the true model.

```{r}
sim_data_1 = data.frame(x_1, x_2, x_3, x_4, x_5, x_6, x_7, x_8, x_9, x_10,
  y = beta_0 + beta_1 * x_1 + beta_2 * x_2 + beta_3 * x_3 + beta_4 * x_4 + 
      beta_5 * x_5 + rnorm(n, 0 , sigma)
)
```

We do a quick check to make sure everything looks correct.

```{r}
head(sim_data_1)
```

Now, we fit an incorrect model.

```{r}
fit = lm(y ~ x_1 + x_2 + x_6 + x_7, data = sim_data_1)
coef(fit)
```

Notice, we have coefficients for `x_1`, `x_2`, `x_6`, and `x_7`. This means that `x_6` and `x_7` are false positives, while `x_3`, `x_4`, and `x_5` are false negatives.

To detect the false negatives, use:

```{r}
# which are false negatives?
!(signif %in% names(coef(fit)))
```

To detect the false positives, use:

```{r}
# which are false positives?
names(coef(fit)) %in% not_sig
```

Note that in both cases, you could `sum()` the result to obtain the number of false negatives or positives.

**(a)** Set a seed equal to your birthday; then, using the given data for each `x` variable above in `sim_data_1`, simulate the response variable `y` 300 times. Each time,

- Fit an additive model using each of the `x` variables.
- Perform variable selection using backwards AIC.
- Perform variable selection using backwards BIC.
- Calculate and store the number of false negatives for the models chosen by AIC and BIC.
- Calculate and store the number of false positives for the models chosen by AIC and BIC.

Calculate the rate of false positives and negatives for both AIC and BIC. Compare the rates between the two methods. Arrange your results in a well formatted table.

```{r, results='hide'}
library(knitr)

birthday = 19950419
set.seed(birthday)
num_sims = 300

false_neg_aic = rep(0,num_sims)
false_neg_bic = rep(0,num_sims)
false_pos_aic = rep(0,num_sims)
false_pos_bic = rep(0,num_sims)

for(i in 1:num_sims){
  sim_data_1$y = beta_0 + beta_1 * x_1 + beta_2 * x_2 + beta_3 * x_3 + beta_4 * x_4 + beta_5 * x_5 + rnorm(n, 0 , sigma)
  fit = lm(y ~ ., sim_data_1)
  n = length(resid(fit))
  AIC = step(fit, direction = "backward")
  BIC = step(fit, direction = "backward", k = log(n))
  false_neg_aic[i] = sum(!(signif %in% names(coef(AIC))))
  false_neg_bic[i] = sum(!(signif %in% names(coef(BIC))))
  false_pos_aic[i] = sum(names(coef(AIC)) %in% not_sig)
  false_pos_bic[i] = sum(names(coef(BIC)) %in% not_sig)
}


```

```{r}
t = data.frame(Criteria = c("AIC","BIC"), False_Neg = c(sum(false_neg_aic) / num_sims, sum(false_neg_bic) / num_sims), False_Pos = c(sum(false_pos_aic) / num_sims, sum(false_pos_bic) / num_sims ))

kable(t)
```


**(b)** Set a seed equal to your birthday; then, using the given data for each `x` variable below in `sim_data_2`, simulate the response variable `y` 300 times. Each time,

- Fit an additive model using each of the `x` variables.
- Perform variable selection using backwards AIC.
- Perform variable selection using backwards BIC.
- Calculate and store the number of false negatives for the models chosen by AIC and BIC.
- Calculate and store the number of false positives for the models chosen by AIC and BIC.

Calculate the rate of false positives and negatives for both AIC and BIC. Compare the rates between the two methods. Arrange your results in a well formatted table. Also compare to your answers in part **(a)** and suggest a reason for any differences.

```{r}
set.seed(94)
x_1  = runif(n, 0, 10)
x_2  = runif(n, 0, 10)
x_3  = runif(n, 0, 10)
x_4  = runif(n, 0, 10)
x_5  = runif(n, 0, 10)
x_6  = runif(n, 0, 10)
x_7  = runif(n, 0, 10)
x_8  = x_1 + rnorm(n, 0, 0.1)
x_9  = x_1 + rnorm(n, 0, 0.1)
x_10 = x_2 + rnorm(n, 0, 0.1)

sim_data_2 = data.frame(x_1, x_2, x_3, x_4, x_5, x_6, x_7, x_8, x_9, x_10,
  y = beta_0 + beta_1 * x_1 + beta_2 * x_2 + beta_3 * x_3 + beta_4 * x_4 + 
      beta_5 * x_5 + rnorm(n, 0 , sigma)
)
```

```{r, results='hide'}
false_neg_aic_2 = rep(0,num_sims)
false_neg_bic_2 = rep(0,num_sims)
false_pos_aic_2 = rep(0,num_sims)
false_pos_bic_2 = rep(0,num_sims)

for(i in 1:num_sims){
  sim_data_2$y = beta_0 + beta_1 * x_1 + beta_2 * x_2 + beta_3 * x_3 + beta_4 * x_4 +beta_5 * x_5 + rnorm(n, 0 , sigma)
  fit = lm(y ~ ., sim_data_2)
  n = length(resid(fit))
  AIC = step(fit, direction = "backward")
  BIC = step(fit, direction = "backward", k = log(n))
  false_neg_aic_2[i] = sum(!(signif %in% names(coef(AIC))))
  false_neg_bic_2[i] = sum(!(signif %in% names(coef(BIC))))
  false_pos_aic_2[i] = sum(names(coef(AIC)) %in% not_sig)
  false_pos_bic_2[i] = sum(names(coef(BIC)) %in% not_sig)
}


```

```{r}
t = data.frame(Criteria = c("AIC","BIC"), False_Neg = c(sum(false_neg_aic_2) / num_sims, sum(false_neg_bic_2) / num_sims), False_Pos = c(sum(false_pos_aic_2) / n, sum(false_pos_bic_2) / num_sims ))

kable(t)
```


